Synthesizing Quotient Lenses

### 2017-10-03 papers/QuotientLensSynth/main.tex
Added explanation for QREs and also notion of unambiguity with related syntax

q := \; &id(R) \; | \; q \mapsto s \; | \; squash (R, R', f) \; | \;
strings.\\
Each QRE $q$ enables us to express

@@

Observe that since $Canonizer(q)$ computes
$Canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that $w$ is
equivalent to $k$ mod $EqRel(q)$, then the equivalence classes of $EqRel(q)$
are the same as the fibres of $Canonizer(q)$.

@@

@@

  \item $q = id(R)$

@@

  \item $W(id(R)) = R$,
  \item $K(id(R)) = R$, 

@@

For all $w, w' \in W(id(R))$, $w \; EqRel(q) \; w'$ if and only if $w =

@@

$Canonizer(id(R))(r) = r$

@@

\UnaryInfC{$id(R)$ is well formed}

@@

\item $q=perm \; (q_1, \ldots, q_n) \; with \; q$

@@

\AxiomC{$\forall \sigma \neq \theta, \; R_{\sigma(1)} \cdot S \cdot \ldots \cdot
S \cdot R_{\sigma(n)} \cap R_{\theta(1)} \cdot S \cdot \ldots \cdot S \cdot
R_{\theta(n)} = \varnothing$} \AxiomC{$\forall \sigma, \; R_{\sigma(1)} \cdot^!
S \cdot^! \ldots \cdot^! S \cdot^! R_{\sigma(n)}$}
\BinaryInfC{$permute \; (R_1, \ldots, R_n) \; with \;  S$ is well formed}

@@

$$\ell := id(R) \; | \; const(s, s') \; | \;  swap(\ell,

@@

\llbracket id(R) \rrbracket &= \{(r, r) \; | \; r \in \mathcal{L}(R)\}\\

@@

\UnaryInfC{$id(\ell): id(R) \Leftrightarrow id(S)$}

@@

\subsection{Normal Forms for Quotient Lenses in $\mathcal{Q}$}
\begin{theorem}\label{normal form}
If there is a derivation $q : c \Leftrightarrow c'$ in
$\mathcal{Q}$ then there exists a bijective lens $\ell : K(c) \Leftrightarrow
K(c')$ such that
\begin{align*}
\llbracket q \rrbracket.get &= \llbracket \ell \rrbracket\circ Canonizer(c)\\
\llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
Canonizer(c')
\end{align*}
\end{theorem}
\begin{proof}
Assume that $q : c \Leftrightarrow c'$. We proceed by induction over the
derivation $q : c \Leftrightarrow c'$.
\begin{enumerate}
  \item
  $id(\ell): id(R) \Leftrightarrow id(S)$ where $\ell : R \Leftrightarrow
S$. Then
  \begin{align*}
  \llbracket id(\ell) \rrbracket.get &=  \llbracket \ell \rrbracket = \llbracket \ell \rrbracket \circ
  id_{\mathcal{L}(R)} = \llbracket \ell \rrbracket \circ Canonizer(id(R)), \text{ and }\\
  \llbracket id(\ell) \rrbracket.put &= \llbracket \ell \rrbracket^{-1} = \llbracket \ell \rrbracket^{-1} \circ
  id_{\mathcal{L}(S)} = \llbracket \ell \rrbracket^{-1} \circ Canonizer(id(S))
  \end{align*}
  \item
  $lquot(c, q'): c' \circ c \Leftrightarrow c''$ where $q' : c' 
  \Leftrightarrow c''$, $c$ is well formed and $K(c) = W(c')$. Then
\begin{align*}
  \llbracket q \rrbracket.get  &= \llbracket q'
  \rrbracket.get \circ Canonizer(c)\\
  \llbracket q \rrbracket.put &= \llbracket q' \rrbracket.put
  \end{align*}
  By the induction hypothesis, there exists a bijective lens $\ell :
  K(c') \Leftrightarrow K(c'')$ such that 
  \begin{align*}
\llbracket q' \rrbracket.get &= \llbracket \ell \rrbracket \circ Canonizer(c')\\
\llbracket q' \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
Canonize(c'')
\end{align*}
Consequently
\begin{align*}
  \llbracket q \rrbracket.get  &= (\llbracket \ell \rrbracket \circ
  Canonizer(c')) \circ Canonizer(c) = \llbracket \ell \rrbracket \circ
  (Canonizer(c' \circ c))\\
  \llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
  Canonize(c'')
  \end{align*}

>>>>

Quotient Regular Expressions (or QREs) enable us to give richer specifications
for data formats than ordinary regular expressions. More concretely, each
Quotient Regular Expression (or QRE) $q$ enables us to express

@@

  All data to be transformed must match $W(q)$, the ``whole'' regular
  expression of $q$. However, before transforming the using a lens, the data
  will be quotiented out by the equivalence relation $EqRel(q)$ which has $K(q)$
  as a complete set of representatives. That is, for each string $w \in
  W(q)$, there is a unique $k \in K(q)$ such that $w \; EqRel(q) \; k$). This
  unique representative $k$ is chosen using the ``canonizing'' function
  $Canonizer(q)$.\\
  Observe that since $Canonizer(q)$ computes $Canonizer(q)(w)$ as the unique $k$
  in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $EqRel(q)$,
  then the equivalence classes of $EqRel(q)$ are the same as the fibres of
  $Canonizer(q)$.
  \subsection{Syntax of QREs}
q := \; &R \; | \; q \mapsto s \; | \; squash (R, R', f) \; | \;
strings.

@@

\subsubsection{Preliminaries}
The transformations that will be performed on the data will require
that the data is matched against regular expressions. Consequently, we require
that regular expressions used by QREs be \textit{strongly unambiguous}, a
condition which ensures that if a string $s$ matches a regular expression $R$,
then $s$ matches $R$ uniquely.\\
To this end, we say that regular expressions $R$ and $S$ are
\textit{unambiguosly concatenable}, written $R \cdot^! S$ if for all strings
$r, r' \in \mathcal{L}(R)$ and $s, s' \in \mathcal{L}(S)$, if $r \cdot s = r'
\cdot s'$, then $r = r'$ and $s = s'$. We say that a regular expression $R$ is
\textit{unambiguosly iterable}, written $R^{*!}$ if for all strings $r_1,
\ldots, r_m$ and $r'_1, \ldots, r'_n \in \mathcal{L}(R)$, if $r_1 \cdot \ldots
\cdot r_m = r'_1 \cdot \ldots \cdot r'_n$, then $m = n$ and $r_i = r'_i$.\\
We say that a regular expression $R$ is \textit{strongly unambiguous} if and
only if (1) $R = \varnothing$, or (2) $R = S_1 \cdot S_2$ with $S_1, S_2$
strongly unambiguous and $S_1 \cdot^! S_n$, or (3) $R = S_1 \; | \; S_2$ with
$S_1, S_2$ strongly unambiguous and $\mathcal{L}(S_1) \cap \mathcal{L}(S_2) =
\varnothing$, or (4) $R = S^*$ with $S$ strongly unambiguous and $S^{*!}$.\\

@@

  \item $q = R$

@@

  \item $W(R) = R$,
  \item $K(R) = R$, 

@@

For all $w, w' \in W(R)$, $w \; EqRel(q) \; w'$ if and only if $w =

@@

$Canonizer(R)(r) = r$

@@

\UnaryInfC{$R$ is well formed}

@@

\item $q=perm \; (q_1, \ldots, q_n) \; with \; q$. Given regular expressions
$R_1, \ldots, R_n$ and $S$, let $\prod({R_i}, S) = R_1 \cdot S \cdot \ldots
\cdot S \cdot R_n$.

@@

\AxiomC{$\forall \sigma \neq \theta, \; \prod(\{W(q_{\sigma(i)})\}, W(q))\cap
 \prod(\{W(q_{\theta(i)})\}, W(q)) = \varnothing$}
\AxiomC{$\forall \sigma, \; K(q_{\sigma(1)}) \cdot^! K(q) \cdot^! \ldots 
\cdot^! K(q_{\sigma(n)})$}
\BinaryInfC{$permute \; (q_1, \ldots, q_n) \; with \;  q$ is well formed}

@@

$$\ell := R \; | \; const(s, s') \; | \;  swap(\ell,

@@

\llbracket R \rrbracket &= \{(r, r) \; | \; r \in \mathcal{L}(R)\}\\

@@

\UnaryInfC{$id(\ell): R \Leftrightarrow id(S)$}

@@

\subsection{Normal Forms for Quotient Lenses in $\mathcal{Q}$}
\begin{theorem}\label{normal form}
If there is a derivation $q : c \Leftrightarrow c'$ in
$\mathcal{Q}$ then there exists a bijective lens $\ell : K(c) \Leftrightarrow
K(c')$ such that
\begin{align*}
\llbracket q \rrbracket.get &= \llbracket \ell \rrbracket\circ Canonizer(c)\\
\llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
Canonizer(c')
\end{align*}
\end{theorem}
\begin{proof}
Assume that $q : c \Leftrightarrow c'$. We proceed by induction over the
derivation $q : c \Leftrightarrow c'$.
\begin{enumerate}
  \item
  $id(\ell): R \Leftrightarrow id(S)$ where $\ell : R \Leftrightarrow
S$. Then
  \begin{align*}
  \llbracket id(\ell) \rrbracket.get &=  \llbracket \ell \rrbracket = \llbracket \ell \rrbracket \circ
  id_{\mathcal{L}(R)} = \llbracket \ell \rrbracket \circ Canonizer(R), \text{ and }\\
  \llbracket id(\ell) \rrbracket.put &= \llbracket \ell \rrbracket^{-1} = \llbracket \ell \rrbracket^{-1} \circ
  id_{\mathcal{L}(S)} = \llbracket \ell \rrbracket^{-1} \circ Canonizer(id(S))
  \end{align*}
  \item
  $lquot(c, q'): c' \circ c \Leftrightarrow c''$ where $q' : c' 
  \Leftrightarrow c''$, $c$ is well formed and $K(c) = W(c')$. Then
\begin{align*}
  \llbracket q \rrbracket.get  &= \llbracket q'
  \rrbracket.get \circ Canonizer(c)\\
  \llbracket q \rrbracket.put &= \llbracket q' \rrbracket.put
  \end{align*}
  By the induction hypothesis, there exists a bijective lens $\ell :
  K(c') \Leftrightarrow K(c'')$ such that 
  \begin{align*}
\llbracket q' \rrbracket.get &= \llbracket \ell \rrbracket \circ Canonizer(c')\\
\llbracket q' \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
Canonize(c'')
\end{align*}
Consequently
\begin{align*}
  \llbracket q \rrbracket.get  &= (\llbracket \ell \rrbracket \circ
  Canonizer(c')) \circ Canonizer(c) = \llbracket \ell \rrbracket \circ
  (Canonizer(c' \circ c))\\
  \llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
  Canonize(c'')
  \end{align*}
